---
title: "Introduction to R Workshop: RMarkdown Example"
author: "Dr Jeromy Anglim"
output: html_document
---

# Overview of Markdown


# Heading 1 with a hash
## Heading 2 with a double hash
### Heading 3 with a triple hash

You can type normal text and insert line breaks as you like.

To indicate a new paragraph, leave an empty line.

You can include dot points by adding an empty line and then using the
star * symbol.

* Example dot point
* Another example

You can also create numbered dot points.

1. My first point that I want to make
2. My second point that I want to make
3. Another point

You can format text:

* Enclose text in backticks to indicate code: e.g., `mycode`
* Enclose text in double stars for bold: e.g., **bolded text**
* Enclose text in single stars for italics: e.g., *bolded text*

Hyperlinks can be included:

* Put text to display in brackets followed by url in parentheses
  [text to display](http://jeromyanglim.blogspot.com)
* Or you can just put a url: http://jeromyanglim.blogspot.com  

If you forget this, you can access the Markdown Quick Reference in Rstudio.


# RMarkdown
* RMarkdown is a document format that involves interspersing Markdown 
with R Code chunks.

* `knitr` then runs the R code and intersperses the resulting output (e.g., tables, figures, text) in the document.


The easiest way to create an R code chunk is to go to "code - insert chunk" in RStudio.

* R code chunks begin with triple backtick, open brace, r, and then close back tick.
  They can also includes some display options.
* R code chunks end with a triple backtick.


It's generally a good idea to put all your library, data loading, and 
data manipulation, code in the first chunk. 
This allows you to minimise the dependencies between code chunks.

```{r import_libraries, warning=FALSE, message=FALSE}
library(foreign) # read.spss
library(psych) # describe 

cars_raw <- foreign::read.spss('cars.sav', to.data.frame=TRUE)
# get rid of missing data listwise
cars <- na.omit(cars_raw[,c('accel', 'mpg', 'engine', 'horse', 'weight')])
```



# Quick look at data
```{r quick_look}
# What's my sample size
nrow(cars)

# Descriptive statistics
psych::describe(cars)

# Basic correlations
round(cor(cars), 2)
```

# You can also include plots
```{r}
psych::pairs.panels(cars)
```

## Summary of model
```{r}
fit <- lm(accel ~ mpg + engine + horse + weight, data=cars)
# r-square, adjusted r-square, std. error of estimate,
# overall ANOVA, df, p, 
# unstandardised coefficients, sig tests
summary(fit)

```
