---
title: "Introduction to R: Core Language Tutorial"
author: "Dr Jeromy Anglim"
output: pdf_document
---


```{r load_project, message = FALSE, warning = FALSE}
library(ProjectTemplate); load.project()
```             



# Basic Arithmetic and Logical Operations
```{r basic_multiplication}
# You can use R like a calculator
1 + 1 # addition
10 - 9 # subtraction
10 * 10 # multiplcation
100 / 10 # division
10 ^ 2 # exponentiation

abs(-10) # absolute value
ceiling(3.5) # round up to next integer
floor(3.5) # round down to next integer
sqrt(100) # square roots
exp(2) # exponents
pi # mathematical constant pi
exp(1) # mathematical constant e
log(100) # natural logs (i.e., base e)
log(100, base= 10) # base 10 logs


# Use parentheses to clarify order of operations
(1 + 1 ) * 2
1 + (1 * 2)

# You can test for equality
# TRUE and FALSE are keywords
# T and F are synonyms, but are generally discouraged
TRUE
FALSE
1 == 2 # Equality (Return TRUE if equal)
1 != 2 # Inequality (Return FALSE if unequal)
10 > 9 # Greater than
9 < 10 # Less than
10 <= 10 # Less than or equal
2 %in% c(1, 2 ,3) # is the number in the vector

# TRUE and FALSE coerces to 1 and 0 respectively
as.numeric(TRUE)
as.numeric(FALSE)

# Logical converting to 0, 1 is useful
x <- c(2, 5 ,7 ,10, 15)
x > 5
sum(x > 5)  # sum of a 0-1 variable is a count
mean(x > 5) # mean of a 0-1 variable is a proportion
```



# Basic language features
```{r basic language features, error = TRUE}
########################################################
# Assignment:
# To assign values to a variable either use <- or =
# <- is the more common convention in R
x <- 1 + 1
x

# = is the common assignment operator in other programming 
# languages. It does work in R, but is not the convention.
y = 1 + 1
y

########################################################
# Variable name rules:
# Variable names generally
# 1. Start with a letter (lower or uppercase)
# 2. Followed by letters, numbers, underscore (_), or period (.)
# 3. No spaces

# These do not work
# my variable <- 1234
# 1234variable <- 1234
# 1234 <- 1234

# This works
myvariable <- 1234
my_variable <- 1234
my_variable <- 1234
myvariable123 <- 1234
myVariable <- 1234
my.variable <- 1234

# R has many naming conventions
# As a matter of preference, style, and convenience, I prefer:
# 1. Short but descriptive names
#    * Less than 8 characters for names of lists and data.frames
#    * Less than 15 characters for variables names in data.frames
# 2. Use underscore to separate words within a variable name
# 3. Avoid upper case letters

# Names starting with a period are hidden
.myvariable <- 1234
ls()
ls(all.names = TRUE)


########################################################
# Comments:
# Comments are any text on a line following a hash #
# 1. They often appear as the first character of a line
#    to present a whole line comment
# 2. At the end of a common on a line
mean(c(1,2,3,4)) # Example of end of line comment

# 3. Half way through a command at the end of a line
c(1, # Example comment
  2,3,   # Another comment
  4)
  

########################################################
# Spaces: 
# R will generally permits zero, one or more spaces between
# variables, operators, and other syntactic elements. 
# However, appropriate and consistent spacing improves
# the readability of you scripts.
# See Hadley Wickham's style guide:
# http://adv-r.had.co.nz/Style.html

# This is bad but works
x<-c(1,2,3,400)*2
x<-    c (   1,2,3,      400)*   2

# This is more readable:
# Add spaces after variables, operators, commas
x <- c(1, 2, 3, 400) * 2


######################################################
# Multipline line commands
# Commands can generally span multiple lines
# as long as R does not think the command has finished 

# This works
x <- c("apple", 
       "banana") 
x
y <- 10 + 
    10  #this works
y

# This does not work
y <- 10
    + 10 
y


########################################################
# Multiple commands on one line
# You can include more than one command on one line 
# by separat the commands by a semicolon.
# But generally, you should avoid doing this as it is not
# very readable.
x <- c(1, 2); y  <- c(3, 4); z <- rnorm(10)
x;y;z


########################################################
# # R is case sensitive
test <- "lower case"
TEST <- "upper case"
TEST 
test # The original value was not lost
     # because test is different to TEST
Test # This variable does not exist

Test <- "title case"
Test


# tip: It's often simpler to make variables lower case
#      so that you don't have to think about case.
```


# Understanding directories
```{r directories}
# R has a working directory.
# This is important when loading and saving files to disk
getwd() # show the current working directory

# you can use setwd to change the working directory
# setwd("~/blah/myproject")

# Tip: Open RStudio with the Rproj file then the working directory
# will be the directory containing the Rproj file.

# Tips: 
# * Try to avoid spaces in file names 
#   (use hyphen or underscore instead)
# * If on Windows, then disable "hide extensions of 
#   known file types" (see folder options )
# * If you do use spaces, then you'll need to escape the space with
#   a slash  (e.g., ("my\ documents")
# * Use backslash as the directory separator 
# * Store all relevant files for a project within 
#   the project working directory
```

# The Workspace
```{r workspace}
###############################################
# R Sessions:

# Quitting R
# You can end the R Session using the q function
# q() 

# But if you are in Rstudio, it is simpler to:
# * Just quit RStudio and this will quit the R session
# * Use the session menu in RStudio to Restart or Terminate
#   an R session


################################################
# Workspaces and environments:
# list environments 
search()

# Create some objects in the global environment
x <- 1:10
y <- 1:20
data(mtcars) # Add a built-in datset mtcars

# Show objects in the global environment
ls() 

# or look at the environment pane in RStudio

###############################################
# Removing objects:
# Removing named objects with the rm function
rm(x)
ls()
rm(y, mtcars)

# Remove all objects from global workspace
# Option 1. Use the following command
rm(list = ls())
# Option 2. Click the broom object in RStudio Environment pane

##############################################
# Saving objects
# Save all objects in the workspace
save.image() 

x <- 30
y <- 1:10
# Save specific named objects using save function.
# rdata or RData is the standard file exetnsion.
save(x, y, file = "output/y.rdata") 

# Let's remove x and change y
rm(x) 
y <- "changed"
y

# load variables stored in rdata file 
load(file = "output/y.rdata")
x
y

# Tips: 
# * Try to avoid using save.image() to store temporary calculations
# * Instead, try to write scripts that can be run to return you to 
#   your current state of analyses.
```

 
# Data types: Logical, character, numeric
```{r data types}
#######################################################
# Basic data types
# The most common basic vector types are
x <- c(FALSE, TRUE) # logical vector
y <- c("a", "b", "cat", "dog") # character vector
z1 <- c(100, 1, 2, 3) # numeric integer vector
z2 <- c(100.2, 0.4, 0.9) # numeric real/double vector
class(x);  typeof(x); mode(x)
class(y);  typeof(y); mode(y)
class(z1);  typeof(z1); mode(z1)
class(z2);  typeof(z2); mode(z2)

# Checking type of object
# there are a range of "is." functions for that return TRUE
# if object is of corresponding type
# apropos("^is\\.")
is.logical(c(TRUE, TRUE))
is.numeric(c("a", "b"))
is.character(c(1, 2, 3))

#############################################################
# Conversion of Types:
# R has functions that explicitly convert data types
# apropos("^as\\.")
as.character(c(1, 2, 3, 4))
as.numeric(c("1", "2a", "3", "four"))
as.numeric(c(FALSE, FALSE, TRUE, TRUE))

# R often performs conversions implicitly
sum(c(FALSE, TRUE, TRUE)) # converts logical to 0, 1 numeric
paste0("v", c(1, 2, 3)) # converts numeric vector to character
```

# Basic data structures: Vectors, Matrices, Lists, Data.frames
```{r basic data structures}
###########################################################
# Vectors:
# In R, a single value (scalar) is a vector.
x <- 1 # I.e., x is a vector of length 1


# In addition to importing data,
# R has various functions for creating vectors.
c(1, 2, 3, 4) # c stands for combine
1:10 # create an integer sequence 1 to 10
seq(1, 10) # alternative way of creating a sequence 
seq(1, 10, by = 2) # The function has additional options
rep(1, 5) # repeat a value a certain number of times
rep(c(1,2,3), 5) # repeat a value a certain number of times

# as well as many simulation functions which we'll cover later
# Initial examples:
# Sample 10 items with replacement from 
sample(x = c("happy", "funny", "silly"), size = 10, replace = TRUE)

# Sample 20 values from a normal distribution
y <- rnorm(n = 20, mean = 0, sd = 1) 
hist(y) # show values in histogram


# Vectors can have names
x <- c(1,2,3,4,5)
names(x) <- c("a", "b", "c", "d", "e")
x

# Extracting vectors
x[c(1,2)] # by numeric position
x[x < 3] # by logical vector
x[c("b", "c")] # by name




###########################################################
# Matrices:
# All data must be of same type (e.g., numeric, character, logical)
y <- matrix(c(1, 2, 
             4, 5, 
             7, 8 ),
           byrow = TRUE, ncol = 2)
y

class(y)

# number of rows and columns
dim(y)  # Number of rows and columns
nrow(y) # Number of rows
ncol(y) # Number of columns

# Rows and columns can be given names
rownames(y) <- c("a", "b", "c")
colnames(y) <- c("col1", "col2")

# Rows and columns can be indexed
y["a", ] # By rowname
y[, "col1"] # By column name 
y["a", "col1"] # By both

y[c(1,2), ] # By row position
y[,1] # By column position
y[c(2,3), 2] # By column position

###########################################################
# Lists
# Store arbitrary structures of one or more named elements.
# Elements can be of different lengths
# Lists can contain lists can be nested to create tree like structures
# Lists are commonly used for representing results of analyses

w <- list(apple = c("a", "b", "c"), 
          banana = c(1,2), 
          carrot = FALSE,
          animals = list(dog = c("dog1", "dog2"), 
                         cat = c(TRUE, FALSE)))

class(w)

# Accessing one element of list
w$apple # using dollar notation
w[[1]] # by position
w[["apple"]] # by name (double brackets)

# Accessing subset of list
w[c(1, 2)]  # by position (single bracket)
w[c("apple", "banana")] # by name
w[c(FALSE, FALSE, TRUE, TRUE)] # by logical vector

# Quick illustration of a list object returned by
# a statistical function

# We'll simulate some data for two hypothetical groups x and y
# and perform an independent samples t-test.
x <- rnorm(10, mean = 0, sd = 1) 
y <- rnorm(10, mean = 1, sd = 1)
fit <- t.test(x, y)  

# The function 
class(fit) # class does not say list, but it is a list
mode(fit)
str(fit) # show structure of object
names(fit) # show names of elements
# we can view particular elements
fit$statistic 
fit$parameter
fit$p.value
# or extract subsets of the list
fit[c("statistic", "parameter", "p.value")]


###########################################################
# Data Frames:
# Data frames are the standard data strucure used for storing
# data. If you have used other software (e.g., SPSS, Excel, etc.), 
# this is what you may think of as a "dataset".
# Columns can be of different data types (e.g., character, numeric, logical, etc.)
z <- data.frame(var1 = 1:9, var2 = letters[1:9])
z

# Tip: Some functions work with matrices, 
#   some work with data.frames, 
#   and some work with both.
# * If you are wanting to store data like you might store in 
#   a database, then you'll generaly want a data.frame.
# * If you are dealing with a mathematical object that you
#   you want to perform a mathematical operation on, then you generally
#   want a matrix (e.g., correlation matrix, covariance matrix, 
#   distance matrix in MDS, matrices used for matrix algebra).
```





# Working with data frames
```{r exploring data frames}
# Let's use the built-in survey data.frame dataset
library(MASS)
data(survey)
?survey
mydata <- na.omit(survey) # for simplicity I'll exclude missing data
shortdata <- mydata[1:6, 1:5] 
shortdata

################################################
# Extracting observations (i.e., rows) and 
# variables (i.e., columns).
# There are similarities to matrices and lists
# Select observations
shortdata[1:5, ] # by row number
shortdata[c(5,4,3,2,1), ] # re-order
shortdata[ shortdata$Sex == "Female", ] # by logical vector
shortdata[c("1", "2"), ] # by rownames

# Select variables 
shortdata[, c(1,2)] # by position like a matrix
shortdata[c(1,2)] # by position like a list
shortdata[ ,c("Sex", "Fold")] # by name like a matrix
shortdata[c("Sex", "Fold")] #
shortdata$Sex # by name to get a single variable


#################################################
# Names
names(shortdata) # get variable names
colnames(shortdata) # but this also works
rownames(shortdata) # rows can also have names

# Tip: Avoid row names.
# Add another variable to the data.frame to store this information.


# Examine first few rows
head(mydata) # first 6 rows
head(mydata, n = 10) # first 7 rows
tail(mydata) # last few rows
# View(mydata) # Rstudio function to open data in viewer
# or click on the icon in the Environment pane

# How many rows and columns?
dim(mydata) # rows and column counts
nrow(mydata) # row count
ncol(mydata) # column count

# Examine structure
str(mydata)
```

# Getting help
```{r getting help}
# Use question mark (i.e., ?) followed by command name
# to lookup specific command
?mean 
help(mean) # or use help function

# to look up package
help(package = "MASS")

# Press F1 in RStudio on the command name
# mean 

# Use double question mark to do a full-text search on R help
??"factor analysis"

# Search google
# e.g., how to get the mean of a vector using r

# Ask question on Stackoverflow with the R tag
# http://stackoverflow.com/questions/tagged/r
```

# Exercise 1 
```{r Exercises 1}
# 1. Working with vectors
# 1.1 Create a variable called x with 10 values drawn from a 
#    normal distribution (see rnorm)

# 1.2 Use the sum and > operator to work out how many values in x
#    are larger than 1


# 3. Using the cats dataset in the MASS package
library(MASS)
data(cats)
# 3.1 Look up the help file on cats

# 3.2 How many observations are there?

# 3.3 Show the first 10 rows of the cats data.frame

# 3.4 Show the structure of cats using the str function

# 3.5 Extract the female cats and assign to variable fcats

# 3.6 How many rows is in fcats?

```

# Answers 1
```{r Answers for Exercises 1}
# 1. Working with vectors
# 1.1 Create a variable called x with 10 values drawn from a 
#    normal distribution (see rnorm)
x <- rnorm(10)

# 1.2 Use the sum and > operator to work out how many values in x
#    are larger than 1
sum(x > 1)

# 3. Using the cats dataset in the MASS package
library(MASS)
data(cats)
# 3.1 Look up the help file on cats
?cats

# 3.2 How many observations are there?
nrow(cats)

# 3.3 Show the first 10 rows of the cats data.frame
head(cats, 10)

# 3.4 Show the structure of cats using the str function
str(cats)

# 3.5 Extract the female cats and assign to variable fcats
fcats <- cats[ cats$Sex == "F", ]

# 3.6 How many rows is in fcats?
nrow(fcats)

```




# Packages
```{r packages, error = TRUE}
# R has many additional packages 
# To use a package it needs to be installed. 
# You only need to install a package once.
# To use a package, you need to load the package each time
# you use R.


##############################################################
# Installation
# Option 1. Use the install.packages function.
# install.packages("psych")
# Note that some packages rely on other packages.
# depencies = TRUE ensures that dependencies are also installed.
# install.packages("psych", dependencies = TRUE)

# Option 2. Use the package tab in R Studio
# Click install and enter package details





##############################################################
# Loading an installed package
# Option 1. Use the library function
library(psych) # I.e., put this at the start of your script

# Other options 
# 2. We'll talk about ProjectTemplate later
#
# 3. Put it in your R startup file 
#    (not recommended as it reduces reproducibility)

##############################################################
# Common errors
# Not having a package installed is a common error
# If you try to load a package that is not installed.
# e.g., 
library(foo)
# You will get an error
# Error in library(foo) : there is no package called ‘foo’
# This means:
# 1. You mistyped the name of the package, or
# 2. You need to install the pakcage
#     install.packages("foo")
# Note foo is just an example. There is no package called foo.

# Not having a package loaded is a common error
# E.g., there is a function you want to use
detach(package:psych) # used for example to ensure psych is not attached 
# say we wanted to use the fisherz function from the psych package
# but the psych package is not loaded
fisherz(.3)
# We get the error:
#   "Error: object 'fisherz' not found"
# Thus we need to run
library(psych)
fisherz(.3)


##############################################################
# Packages contain additional functions.
# Once the package is loaded, functions are added to the workspace
# list workspace
search()

# To make it clear that a function comes from a particular package
# or to overcome the issue where two packages have functions with the same names
# use double colon (i.e., package::function).
# RStudio also permits auto-completion of function names.
# psych::alpha() # alpha is a funtion in the psych package

```



# Missing data
```{r missing data}
# Missing data is represented in R by NA
x <- c(1, 2, NA, 4)
y <- c("a", "b", NA, "c")
x
y

# To see whether a value is missing
is.na(x)

# If you have missing data, some functions will return NA by default
# rather than returning a value
mean(x)
sd(x)

# Many functions have a na.rm argument
mean(x, na.rm=TRUE)
sd(x, na.rm=TRUE)

# or you remove the missing data
na.omit(x)
mean(na.omit(x))

# na.omit also works on data frames performing listwise deletion
head(survey)
dim(survey)
cleaned_survey <-na.omit(survey)
dim(cleaned_survey)
```

# Getting summaries of data frames
```{r data.frame summaries, warning = FALSE}
library(MASS) # user survey data from MASS package
data(survey) # load an internal dataset
mydata <- survey

# Variable Names
names(mydata)

# Show structure
str(mydata)

# Useful summary of numeric and categorical variables
Hmisc::describe(mydata)

# Common univariate statistics for numeric variables
psych::describe(mydata)
summary(mydata)
```

# Summaries of numeric vectors (or data frame variables)
```{r summaries of numeric vectors}
x <- c(1, 2, 3, 4,5)

# Total
sum(x) # sum of vector
prod(x) # product of vector

# Central tendency
mean(x) # mean of vector
median(x) # median of vector
length(x) # length of vector

# Spread
sd(x) # standard deviation
var(x) # variance
range(x) # min and max of vector
min(x) # minimum of vector
max(x) # max of vector

# Other distributional features
psych::skew(x) # skewness 
psych::kurtosi(x) # kurtosis
```

```{r vector operations}
dat <- data.frame(x = c(1, 2, 3, 4, 5), 
                  y = c(0, 0, 1, 1, 1))
dat

# Vector operations typically operate element wise
dat$z <-  dat$x + dat$y
dat

dat$z <-  dat$x * dat$y
dat

# A single value is recyled through the vector
dat$z <- dat$x + 10
dat
```

# Exercise 2 - Data.frames
```{r exercise 2}
# For this exercise will use the GSS7402 dataset
library(AER)
help(package = AER)
data("GSS7402")
?GSS7402 # to learn about the dataset
# It might be easier to work with a shorter variable name 
gss <- GSS7402

# 1. List the variable names in the gss dataset

# 2. Show the first few rows (hint: the head) of the dataset?

# 3. How many cases are there?

# 4. What is the mean, sd, and range age of the sample

# 5. Use the psych and Hmisc describe functions to describe the samples

# 6. Extract a data.frame with only people over the age of 80

# 7. Get the mean number of children ("kids") for participants 
#    over the age of 80

# 8. Use the mean function to get the mean age at first birth.
#    Hint: there is missing data.
```


# Answers Exercise 2 - Data.frames
```{r answers for exercise 2, warning = FALSE}
# For this exercise will use the GSS7402 dataset
library(AER)
help(package = AER)
data("GSS7402")
?GSS7402 # to learn about the dataset
# It might be easier to work with a shorter variable name 
gss <- GSS7402

# 1. List the variable names in the gss dataset
names(gss)

# 2. Show the first few rows (hint: the head) of the dataset?
head(gss)

# 3. How many cases are there?
nrow(gss)

# 4. What is the mean, sd, and range age of the sample
mean(gss$age)
sd(gss$age)
range(gss$age)

# 5. Use the psych and Hmisc describe functions to describe the samples
psych::describe(gss)
Hmisc::describe(gss)

# 6. Extract a data.frame with only people over the age of 80
gss_over80 <- gss[ gss$age > 80, ]


# 7. Get the mean number of children ("kids") for participants 
#    over the age of 80
mean(gss[ gss$age > 80, "kids"])

# 8. Use the mean function to get the mean age at first birth.
#    Hint: there is missing data.
mean(gss$agefirstbirth) # doesn't work because there is missing data
mean(gss$agefirstbirth, na.rm = TRUE) # doesn't work because there is missing data
```


# String functions
```{r string functions}
paste("hello", "how", "are", "You") # defaults to space separator
paste0("hello", "how", "are", "You") # no separator
paste("apple", "banana", "carrot", "date", sep =", ") # specify arbitrary separator
paste0("v", 1:10) # paticularly useful with vectors

# Extract substring
substr("abcdefghijklmnop", 4, 6)

# Change case
toupper("abcd") # make upper case
tolower("ABCD") # make lower case

mystring <- c("apple", "banana", "carrot", "date", "egg", "fig")
# Identify which strings match a pattern
grep("a", mystring) # index of objects with "a"
grep("a", mystring, value = TRUE) # value of objects with "a"

# get count of number of characters
nchar(mystring)
data.frame(mystring, nchar(mystring))

# Substitute a mystringreplacement text that matches a pattern
questions <- c("How are you?", "What is going on?")
gsub(" ", "_", questions) # replace space with underscore


# R  string manipulation tools are very powerful
# For more information see
?grep
?"regular expression"


# see also Hadley Wickham's package for string manipulation
# It attempts to introduce greater consistency in notation.
# install.packages("stringr")
library(stringr)
help(package = "stringr")
# all functions begin with str_
str_length(mystring) # see nchar
str_sub(mystring, start = 1, end = 3)


# writing output to the console
cat("Hello World!")

# Tab is \t and new line is \n
cat("Hello\t World\nSome more text")

```


# Importing data
```{r reading data}
# A simple option is to export data from your external data
# in csv format and then import the data using csv
# csv
medals <- read.csv("data/practice/medals.csv")
head(medals)
tail(medals)
dim(medals)

# Other delimited formats
medals <- read.table("data/practice/medals.tsv", sep ="\t")

# Read Excel: 
# Read xls files using xls
# Requires that Perl is installed and on the path
# You may need to install Perl if on Windows
# https://www.perl.org/get.html
# library(gdata) 
# medals <- gdata::read.xls("data/practice/medals.xls")

# requires Java
# library(xlsx)
# x <- xlsx::read.xlsx("data/practice/medals.xlsx", sheetIndex = 1)

# More recent package that has no dependencies on external packages 
# readxl
library(readxl)
medals <- readxl::read_excel("data/practice/medals.xls")
medals <- readxl::read_excel("data/practice/medals.xlsx")


# SPSS
library(foreign)
cas <- foreign::read.spss("data/practice/cas.sav", to.data.frame = TRUE)
attr(cas, "variable.labels")
# tip: You may need to think about value labels in your SPSS file
# Specifically, if you have numeric variables that have variable labels, you may
# want to remove the value labels in SPSS or 

# import stata
?read.dta
?read.sas


# General purpose packages 
# The haven package also can read and write SPSS, SAS, and Stata files

# rio package: General purpose import and export
# General purpose import and export tools
# It's a fairly new package so there may still be a few bugs.
# https://github.com/leeper/rio
# github version is currently a little bit ahead of the CRAN version
library(rio)
cas <- rio::import(file="data/practice/cas.sav")
rio::export(cas, file="output/cas.xlsx")
medals <- rio::import(file="data/practice/medals.csv")

# Use ProjectTemplate to auto-import (see discussion later)
```

# Exporting data 
```{r exporting data}
mydata <- data.frame(a = c(1,2,3), b = c("a", "b", "c"))

# Interal R format
# Good option if you need to re-open data in R
save(mydata, file="output/mydata.rdata")
#  load("output/mydata.rdata")


# csv
# Good option if you need to get data into other software
# This should open in almost all other software (e.g. Excel, SPSS, etc.)
write.csv(mydata, file = "output/mydata.csv")
write.csv(mydata, file = "output/mydata-2.csv", row.names = FALSE) # exclude row.names

# If you need more flexibility in terms of delimiters, etc.
write.table(mydata, file = "output/mydata.tsv", sep = "\t") # e.g., tab delimiter

# Exporting to other formats
# There are a range of options for exporting to other formats
# Functionality is often spread around
# Given that the csv option is usually sufficient
library(foreign)
?foreign::write.foreign  # options for exporting to SAS, SPSS, and Stata directly

?rio::export
```

# Exercise 3
```{r Exercise 3}
# 1. Open medals.csv in the data/practice/ directory 
#    and assign to variable medals

# 2. Check that the file imported correctled
#    (a) look at the first few rows, 
#    (b) look at the last few rows,
#    (b) check the structure (i.e., str),
#    (c) Use the Hmisc describe function to check basic properties

# 3. Create a new variable in medals that indicates
#    whether the medals was Gold (TRUE) or Silver/Bronze (FALSE)
#    and call it isgold

# 4. How the number of sum of gold medals 

# 5. Export the medals data.frame to the output folder
#    (a) as a csv file
#    (b) as a native rdata file

# 6. Remove the medals dataset from the workspace 
#    and then load it again from the csv file.
#    Check that it imported correctly. 
# Then remove medals and repeat for the rdata file

```

# Answers for Exercise 3
```{r Answers for Exercise 3}
# 1. Open medals.csv in the data/practice/ directory 
#    and assign to variable medals
medals <- read.csv("data/practice/medals.csv")

# 2. Check that the file imported correctled
#    (a) look at the first few rows, 
#    (b) look at the last few rows,
#    (b) check the structure (i.e., str),
#    (c) Use the Hmisc describe function to check basic properties
head(medals)
tail(medals)
str(medals)
Hmisc::describe(medals)

# 3. Create a new variable in medals that indicates
#    whether the medals was Gold (TRUE) or Silver/Bronze (FALSE)
#    and call it isgold
medals$isgold <- medals$Medal == "Gold"

# 4. How the number of sum of gold medals 
sum(medals$isgold)

# 5. Export the medals data.frame to the output folder
#    (a) as a csv file
#    (b) as a native rdata file
write.csv(medals, "output/medals.csv")
# or technically you may want to do
write.csv(medals, "output/medals.csv", row.names = FALSE)
save(medals, file = "output/medals.rdata")

# 6. Remove the medals dataset from the workspace 
#    and then load it again from the csv file.
#    Check that it imported correctly. 
# Then remove medals and repeat for the rdata file
rm(medals)
medals <- read.csv("output/medals.csv")
head(medals)

rm(medals)
load("output/medals.rdata")
head(medals)
```


# Random variables and distributions
```{r random variables and distributions}
# In statistics, we often want to generate random data with certain properties
# or looking up features of statistical distributions.
# See the following help for list of common distributions is base R
?Distributions

# and see http://cran.r-project.org/web/views/Distributions.html for many more distributions

# Each distribution has four functions that differ in terms of the first letter
# For example, for the normal distribution, you have
dnorm(1) # Density of the value 1 of a standard normal distribution
pnorm(1) # Cumulative distribution function for value of 1 on standard normal distribution
qnorm(.975) # Inverse cumulative distribution function for value of .975
rnorm(5) # Generate 5 random draws from normal distribution

# 
dunif(1) # Density of the value 1 of a uniform distribution (0, 1)
punif(.5) # Cumulative distribution function for value of 1 on uniform distribution
qunif(.975) # Inverse cumulative distribution function for value of .975
runif(5) # Generate 5 random draws from uniform distribution


# Distributions have parameters that can be specified
x <- rnorm(10, mean = 10, sd = 1) # draw 10 from mean of 10
y <- rnorm(10, mean = 11, sd = 1) # draw 10 from mean of 11
dat <- data.frame(x=x, y=y)
dat
boxplot(dat)
```


# Functions
```{r functions}
# You can write functions and these are generally the same as
# the functions you use in R

# For example, I could create a function that printed some text
print_some_text <- function(x = "Hello World") {
    print(x)
}

# If I run the above command, I can then use it
print_some_text() # using the default argument
print_some_text("blah blah blah")  # or to print some other text

# Anatomy of a function 
# Functions have a name
# They take one or more arguments
# Arguments may have default values


# Let's take a more interesting example: Power analysis
# The following data simulates data for two groups and
# examines whether there is a significant difference at .05
# It repeats the process 1000 times and calculates the
# proportion of times it is statistically significant
# (i.e., simluation estimate of the statistical power)

significant <- NULL
for (i in 1:1000) {
    x <- rnorm(30, mean = 0.0, sd = 1) 
    y <- rnorm(30, mean = 0.3, sd = 1) 
    fit <- t.test(x, y)
    fit
    significant[i] <- (fit$p.value < .05)
}
statistical_power <- mean(significant)
statistical_power

# we could convert this to a function
power_group_dif1 <- function() {
    significant <- NULL
    for (i in 1:1000) {
        x <- rnorm(30, mean = 0.0, sd = 1) 
        y <- rnorm(30, mean = 0.3, sd = 1) 
        fit <- t.test(x, y)
        fit
        significant[i] <- (fit$p.value < .05)
    }
    statistical_power <- mean(significant)
    statistical_power
}

power_group_dif1()

# but the beauty of function is that they can make things general
# Let's make the mean of group 2 an argument that can be specified
power_group_dif2 <- function(mean2 = 0.3) {
    significant <- NULL
    for (i in 1:1000) {
        x <- rnorm(30, mean = 0.0, sd = 1)
        y <- rnorm(30, mean = mean2, sd = 1) 
        fit <- t.test(x, y)
        fit
        significant[i] <- (fit$p.value < .05)
    }
    statistical_power <- mean(significant)
    statistical_power
}

# now we can specify different values
power_group_dif2(0) 
power_group_dif2(.3) 
power_group_dif2(.5) 
power_group_dif2(.8) 
power_group_dif2(1) 

settings <- seq(from = 0, to = 2, by = .1)
results <- data.frame(mean2= settings)
results$power <- sapply(results$mean2, function(X) power_group_dif2(X))
plot(results, type = "b")


# obviously it could be made a whole lot more general
power_group_dif3 <- function(mean1 = 0, mean2 = 0.3, sd1 = 1, sd2 = 1, 
                             n1 = 30 , n2 = 30, ksimulations = 1000,
                             alpha_criterion = .05) {
    significant <- NULL
    for (i in 1:ksimulations) {
        x <- rnorm(30, mean = mean1, sd = sd1)
        y <- rnorm(30, mean = mean2, sd = sd2) 
        fit <- t.test(x, y)
        fit
        significant[i] <- (fit$p.value < alpha_criterion)
    }
    statistical_power <- mean(significant)
    statistical_power
}

power_group_dif3(mean1 = 10, mean2 = 11, sd1 = 1, sd2 = 1, 
                             n1 = 100 , n2 = 100, ksimulations = 1000,
                             alpha_criterion = .01)

```

# Debugging functions
```{r debugging functions}
# debugging functions
print_some_text <- function(x = "Hello World") {
    print(x)
}
debugonce(print_some_text) # activates debugging on the function
print_some_text()

# many other useful functions
?traceback # provide further information when an error occurs
?browser # place in function 
```


# Viewing source code for internal functions
```{r viewing source code, error = TRUE, results = "hide"}
# Option 1: type function name
t.test
cor
power.t.test


# Option 2: 
# S3 Methods
# Some functions are generic and operate differently depending
# on the class of the first argument
# mean
# print
# summary

#  Methods will list the actual function names called
methods(mean)
methods(print)
methods(summary)

mean.default
summary.table


# Option 3:
# Some functions are part of packages but are not exported
# I.e., they are intended for internal use, but
# they are often quite useful
library(ProjectTemplate)

# Double colon shows the functions exported from a package
# i.e., packagename::function
ProjectTemplate::run.project
    
# Triple colon shows internal functions
# i.e., packagename:::function
ProjectTemplate:::xls.reader

# Also, see the getAnywhere function
xls.reader # this doesn't work
getAnywhere(xls.reader) # this does work
```


# Exercise 4
```{r Exercise 4}
library(MASS)
data(mammals)
?mammals
head(mammals)
# 1. Create a function that takes a single argument x
#    and prints that value twice.
#    use the function to print "hello world" twice


# 2. Divide mammall brain weight (g) by body weight (kg) and 
#    get the mean of this value


# 3. Write a function that takes arguments x and y
#    and returns the mean of x divided by y


# 4. Apply the function to get the mean ratio of brain to body size


# 5. Modify the ratio function to return a list with 
#    (a) the mean of x divided by y, and
#    (b) the sd of x divided by y. 
#    Then apply to mammals data as above.


# 6. Step through the code for the correlation function 


# 7. Show the source code for 
#    (a) the t.test function, 
#    (b) the summary method for lm objects
#    (c) the alpha function in the psych package

```

# Answers 4
```{r Answers for Exercise 4}
library(MASS)
data(mammals)
?mammals
head(mammals)
# 1. Create a function that takes a single argument x
#    and prints that value twice.
#    use the function to print "hello world" twice
print_twice <- function(x) {
    print(x)
    print(x)
}
print_twice("hello world")


# 2. Divide mammall brain weight (g) by body weight (kg) and 
#    get the mean of this value
mean(mammals$brain / mammals$body )

# 3. Write a function that takes arguments x and y
#    and returns the mean of x divided by y
mean_ratio <- function(x, y) {
    mean(x / y)
}

# 4. Apply the function to get the mean ratio of brain to body size
mean_ratio(mammals$brain, mammals$body)

# 5. Modify the ratio function to return a list with 
#    (a) the mean of x divided by y, and
#    (b) the sd of x divided by y. 
#    Then apply to mammals data as above.
mean_ratio <- function(x, y) {
    ratioxy <- x / y
    list(mean_ratio = mean(ratioxy),
         sd_ratio = sd(ratioxy))
}

# 6. Step through the code for the correlation function 
# debugonce(cor)
cor(mammals$brain, mammals$body, method = "spearman")


# 7. Show the source code for 
#    (a) the t.test function, 
#    (b) the summary method for lm objects
#    (c) the alpha function in the psych package
# t.test
# summary.lm
# psych::alpha
```







